---
title: "Homework 2 P3"
author: "Amy Ly"
date: "2/9/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Context

The data below are launch temperatures (degrees Farenheit) and an indicator o O-ring failures for 24 space shuttle launches prior to the space shuttle Challenger disaster of January 27, 1986. Note that 0 indicates "No Failure" while 1 indicates "Failure"

```{r}
data <-
  data.frame(temperature = c(53, 56, 57, 63, 66, 67, 67, 67, 
                             68, 69, 70, 70, 70, 70, 72, 73, 
                             75, 75, 76, 76, 78, 79, 80, 81), 
                  failure = c(1, 1, 1, 0, 0, 0, 0, 0, 
                              0, 0, 0, 1, 1, 1, 0, 0, 
                              0, 1, 0, 0, 0, 0, 0, 0))

head(data)

```

### Exploratory Data Analysis

```{r}

plot(data$temperature, data$failure/nrow(data), 
     xlab="temperature", ylab="failure rate")
```

Visualize the data

### Part a

Fit the logistic regression of Failure on Temperature. 

The estimated coefficients and their standard errors are then summarized below. 


```{r}

##  One column binary response 

n <- nrow(data)
y0=as.numeric(data$failure)

X=matrix(1,n,2)
X[,2]=data$temperature


#### Fit logistic regressions 

logit.y = glm(y0~data$temperature, family=binomial)

summary(logit.y)

```

### Part b

Test whether the coefficient of Temperature is 0 using Wald's test. Report a one-sided p-value where the alternative hypothesis is that the coefficient is negative (that the odds of failure decreases with increasing temperature).


```{r}

logit.y$coef
V = vcov(logit.y)

#Wald test

Wald <-(logit.y$coef[2]-0)/sqrt(solve(V)[2,2])

Wald

p.val <- pnorm(Wald, lower.tail=TRUE)

p.val

```

Since the resulting p-value is 0.499 and much greater than the $\alpha =0.05$, we do not have enough evidence to say that the coefficient of Temperature is not equal to 0. 

### Part c

Give the 95% confidence interval for the coefficient of Temperature. 

```{r}

#95% CI for beta
c(logit.y$coef[2]-1.96*sqrt(V[2,2]), 
  logit.y$coef[2]+1.96*sqrt(V[2,2]))


```

The 95% confidence interval does not contain 0, which indicates that there is a possibility that the coefficient is negative. 

### Part d

What is the estimated probability of failure at 31 degrees Farenheit?

```{r}
# At temp = 31, parameter theta = alpha + beta*31
theta_hat = logit.y$coef[1]+logit.y$coef[2]*31

exp(theta_hat) # Predicted survival odds 

exp(theta_hat)/(1+exp(theta_hat)) # Predicted survival probability 

```

At 31 degrees Farenheit, the probability of O-ring failure is about 99.6%.